### 2018-07-19 submission/manuscript.tex
update manuscript from full read through

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study,
we focus on the amplification step's contribution to this overall error.
To accomplish this we assessed 16S rRNA gene sequencing results in human
fecal and mock community samples after using different high fidelity
(HiFi) DNA polymerases and number of amplification cycles.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition by the
Bray-Curtis index also was assessed based on polymerase and number of
cycles. Based on these findings individual taxa differences based on
polymerase and number of cycles was investigated. Finally, Random Forest
models were created to test whether the bacterial community was better
at classifying polymerases, number of cycles, or individual donor. We
also assessed whether the most important taxa in the polymerase and
number of cycle Random Forest models were also important in the model
for individual donors.

\textbf{Results.} Predictably, we found noticeable differences in
relative abundance based on high and low GC content (P-value
\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
community samples the variation in the number of OTUs detected by the
polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
When analyzing fecal samples we observed smaller differences in the
number of OTUs. Additionally, the median number of OTUs only varied by
HiFi DNA polymerase used at 35 cycles and not at 20 cycles like the mock
community (P-value \textless{} 0.0001). Random Forest models were most
successful at classifying individual donor samples rather than
polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
Additionally, the most important OTUs in the polymerase and number of
cycle models were not the most important in the individual donor sample
model.

\textbf{Conclusions.} Although there were 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, bacterial
community composition differences are mostly only detectable in mock
communities. Collectively, these results provide evidence that smaller
biological differences between groups, based on 16S rRNA gene sequencing
of fecal samples, can be consistently detected regardless of polymerase
and number of cycles used.

@@

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Some of this variation could be
due to error introduced during the 16S rRNA gene sequencing workflow. A
typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not as well
characterized.

@@

Based on these observations, we created Random Forest models to classify
samples based on number of cycles or polymerase used. When assessing the
Random Forest models there was a difference between models built to
classify number of cycles versus those built to classify polymerase used
{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
was successful at correctly classifying samples based on 16S rRNA gene
sequencing data (cycles model, probability of correct classification =
0.42 (0.4 - 0.43)(min - max) versus polymerase model, probability of
correct classification = 0.39 (0.37 - 0.4)). Our observations in mock
communities suggest that although differences are detected based on
number of cycles or polymerase used, the overall community-wide
differences are subtle.

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
are only small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that polymerases and
number of cycles had no affect on the bacterial community composition in
fecal samples (P-value = 1). There was also no interaction between
individuals, number of cycles, and polymerase used (P-value = 1). When
using PERMANOVA to test for community differences based on any of the
number of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03). Additionally, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 sequences
at 15 cycles. Within each respective 5-cycle increment comparison there
was no difference in Bray-Curtis index between the polymerases (P-value
\textgreater{} 0.05) {[}Figure 4B{]}. However, there was an overall
decrease in Bray-Curtis index by cycle comparison group (e.g.~the 15
cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle
group) (P-value \textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's
post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle
comparison groups had a higher Bray-Curtis index than the 30 cycle
versus 35 cycle comparison group (P-value \textless{} 0.05). In addition
to assessing community composition by Bray-Curtis index, we also
investigated differences in OTUs based on polymerase at each number of
cycle used and no significant differences could be found (P-value
\(\geqslant\) 0.01, corrected P-value = 1). Overall, these results
suggest that in fecal samples the number of cycles and polymerase used
only minimally change the bacterial community composition.

@@

The small differences observed due to error rate and chimera prevalence
may be due to the actual DNA polymerase family being used within the PCR
mix. DNA polymerases from distinct families have different binding
affinities and error correction capacity (Ishino \& Ishino, 2014).
However, based on the observations within our study this is unlikely to
be the only contributor. Within our study the highest and lowest chimera
prevalence both belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the information supplied by the respective
manufacturers, the differences between the two PCR mixtures are not
immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase and a platinum
\emph{Taq} antibody. Since it is not possible to know everything about
the mixture beyond what was willingly provided by the manufacturer, it
is possible that differences in how the recombinant \emph{Taq} was
generated or other compounds within the PCR mixture could be a
contributing factor for the differences in error rate and chimera
prevalence. Beyond the choice of the type of polymerase, there may be
other ways to reduce the affect of polymerase dependent error rates and
chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference
when using the Bray-Curtis index to measure the bacterial community.
Specifically, there was no difference in distance between successive
5-cycle increments within samples between the polymerases when using the
Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
is that many of the OTUs generated by polymerase dependent error rates
and chimera prevalence are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight these OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate some of the observed
polymerase dependent differences in error rate and chimera prevalence.
Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
Vargas, 1996), richness) may be less robust to polymerase dependent
error rates and chimera prevalence. When choosing a distance metric,
careful consideration of the biases introduced from the PCR step of the
16S rRNA gene sequencing workflow need to be taken into account. One
possible way to better choose polymerases may be based on the DNA
polymerase family used in the PCR mixture.

@@

\textbf{\emph{Human and mock samples.}} Fecal samples were obtained from
4 individuals who were part of the Enterics Research Investigational
Network (ERIN). The processing and storage of these samples were
previously published (Seekatz et al., 2016). Other than confirmation
that none of these individuals had a \emph{Clostridium difficle}
infection, clinical data and other types of meta data were not utilized
or accessed for this study. All samples were extracted using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR protocol.}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
polymerases activation time was 2 minutes, unless a different activation
was specified by the manufacturer. The annealing and extension time for
Platinum and Accuprime followed a previously published protocol (Kozich
et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time also were from a
previously published protocol (Gohl et al., 2016). For Phusion, the
company defined activation and annealing times were used while the
extension time followed the Accuprime and Platinum settings.

The number of cycles in the PCR for fecal and mock samples started at 15
and increased by 5 up to 35 cycles, with amplicons used at each 5-step
increase for sequencing. The PCR of fecal DNA samples consisted of all 4
samples at 15, 20, 25, 30, and 35 cycles for each polymerase (total
sample n=100). The mock communities had 4 replicates at 15, 20, 25, and
35 cycles and 10 replicates for 30 cycles for all polymerases (total
samples n=130). No mock community sample had enough PCR product at 15
cycles for adequate 16S rRNA gene sequencing.

\textbf{\emph{Sequence processing.}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016). For
simplicity, the use of the term `error' within the context of this study
moving forward encompasses changes in the downstream sequencing results
due to the amplification step (e.g.~higher GC content genomes, base
substitutions, deletions, insertions, non-specific amplification,
chimeras, etc.).

\textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis. The
analysis of the mock community of each polymerase for GC-based
amplification differences, sequence error rate, number of sequences with
an error, base substitution, and numbers of chimeras before and after
chimera removal with VSEARCH was assessed. Additionally, the correlation
between the number of chimeras and the number of OTUs was also assessed.
The total number of OTUs, taxa differences, and Bray-Curtis indices were
analyzed for both the fecal and mock community samples. Finally, Random
Forest models were created to assess whether classification of
polymerase, number of cycles, or individual could be performed best
using 16S rRNA gene sequencing data. For the model used to classify
individuals, all samples were included from all number of cycles and
polymerases but only the individual labels were used. Both number of
cycles and polymerase models included all samples but only the number of
cycles or polymerase label was used for each respective model.
Additionally, overlap between the most important OTUs to the three
models was assessed using mean decrease in accuracy (MDA).

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the whole genome of the bacterial
species used in the mock community. Differences in the total number of
OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the
comparison of the number of OTUS in fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation. Bray-Curtis distance matrices were generated using
mothur. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and Kruskal-Wallis tests
within R. The Random Forest models were run using the caret package (v
6.0.78) (Jed Wing et al., 2017). A total of 100 10-fold CV runs on
different 80/20 splits of the data was run to generate a range of the
Logloss value. The probability of a correct call was obtained from this
Logloss value by taking the negative natural logarithm. For both error
and chimera analysis, samples were tested using Kruskal-Wallis with a
Dunns post-hoc test. Where applicable, correction for multiple
comparison utilized the Benjamini-Hochberg method (Benjamini \&
Hochberg, 1995).

\textbf{\emph{Reproducible methods.}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found on the SRA at the following accession
number SRP132931.

>>>>

\textbf{Background.} A typical 16S rRNA gene sequencing workflow can be
divided into preservation, extraction, amplification, and sequencing
steps. At each of these stages error can be introduced that will change
the underlying bacterial community composition results. In this study,
we focus on the amplification step's contribution to this overall error.
To accomplish this we assessed 16S rRNA gene sequencing results in mock
community and human fecal samples after using different amplification
cycles and high fidelity DNA polymerases.

\textbf{Methods.} Amplification of mock communities (technical
replicates n=4) consisting of previously isolated whole genomes of 8
different bacteria were amplifeid at 15, 20, 25, 30, and 35 cycles using
Accuprime, Kappa, Phusion, Platinum, or Q5 high fidelity DNA polymerase.
For fecal samples (n=4) we first extracted DNA using a PowerMag DNA
extraction kit with a 10 minute bead beating step and then used the same
approach as the mock community for the PCR amplification step. We first
assessed GC dependent differences, error rate, sequence error
prevalence, chimera prevalence, and correlation between chimera
prevalence and number of Operational Taxonomic Units (OTUs) by
polymerase and number of cycles. Next, differences in the number of OTUs
was examined based on the polymerase and number of cycles used.
Additionally, differences in the bacterial community composition was
assessed based on polymerase and number of cycles using the Bray-Curtis
index. Based on these findings individual taxa differences based on
polymerase and number of cycles was also investigated. Finally, Random
Forest models were created to test whether the bacterial community was
better at classifying polymerases, number of cycles, or individual
donor. We also assessed whether the most important taxa in the
polymerase and number of cycle Random Forest models were also important
in the model for individual donors.

\textbf{Results.} Predictably, we found noticeable differences in
relative abundance based on high and low GC content (P-value
\(\leqslant\) 0.04). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478). When analyzing mock
community samples the variation in the number of OTUs detected by the
polymerases was observable as early as 20 cycles (P-value = 0.002). When
analyzing fecal samples we observed smaller differences in the number of
OTUs between polymerases. The median number of OTUs only varied by high
fidelity DNA polymerase used at 35 cycles and not at 20 cycles like the
mock community (P-value \textless{} 0.0001). Random Forest models were
most successful at classifying individual donor samples rather than
polymerase or number of cycles used (P-value \(\leqslant\) 5.49e-07).
Additionally, the most important OTUs in the polymerase and number of
cycle models were not the most important in the individual donor model.

\textbf{Conclusions.} Although there were 16S rRNA gene sequencing
differences based on polymerase and number of cycles used, bacterial
community composition differences were mostly only detectable in mock
communities. Collectively, these results provide evidence that smaller
biological differences between groups, based on 16S rRNA gene sequencing
of human fecal samples, can be consistently detected regardless of
polymerase and number of cycles used.

@@

The bacterial community is reported to vary between case and control for
a number of diseases (Turnbaugh et al., 2008; Sze et al., 2015; Baxter
et al., 2016; Bonfili et al., 2017). However, for diseases like obesity,
the taxa identified have varied widely depending on the study (Turnbaugh
et al., 2008; Zupancic et al., 2012). Although some of the differences
could be attributed to study design, error introduced during the 16S
rRNA gene sequencing workflow could also be another source of variation.
A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. The preservation
and extraction stages of the 16S rRNA gene sequencing workflow have been
the most extensively studied (Salter et al., 2014; Song et al., 2016;
Bassis et al., 2017; Kim et al., 2017). For preservation and extraction
stages of the workflow, it has been consistently found that there are
errors in the observed bacterial community composition based on the kits
used, but that these differences are smaller than the overall biological
difference measured between samples with different kits (Song et al.,
2016; Bassis et al., 2017). Since these studies use the same PCR
approach while varying preservation or extraction method, the
contribution of PCR error to this overall workflow is not as well
characterized.

@@

Based on these observations, we created Random Forest models to classify
samples based on number of cycles or polymerase used. When assessing the
Random Forest models there was a difference between models built to
classify number of cycles versus those built to classify polymerase used
{[}P-value = 5.06e-07{]}. Although there was a difference, neither model
was overly successful at correctly classifying samples based on 16S rRNA
gene sequencing data (cycles model, probability of correct
classification = 0.42 (0.4 - 0.43)(min - max) versus polymerase model,
probability of correct classification = 0.39 (0.37 - 0.4)). Our
observations in mock communities suggest that although differences are
detected based on number of cycles or polymerase used, the overall
community-wide differences are not as large.

\textbf{\emph{Few differences based on number of cycles and polymerase
used were identified in fecal samples.}} Unlike mock community samples,
an overall difference in the number of OTUs was only detected between
polymerases at 35 cycles in fecal samples (F-stat \textgreater{} 16.35,
P-value = 9.7e-05) {[}Table S4{]}, and a Tukey post-hoc test failed to
identify any differences between specific polymerases used (P-value
\textgreater{} 0.05) {[}Table S5{]}. However, differences in the range
of the number of OTUs detected for fecal samples was dependent on the
polymerase used (e.g.~Accuprime at 35 cycles = 84 - 106 versus Phusion
at 35 cycles = 84 - 136) {[}Figure 4A{]}. There was also a trend for
lower number of cycles (15-20) to result in a reduced range in the
number of OTUs versus higher number of cycles (25, 30, and 35) for all
polymerases (e.g.~Phusion at 15 cycles = 10 - 19 versus Phusion at 35
cycles = 84 - 136) {[}Figure 4A{]}. Based on these observations, there
are small differences in the total number of OTUs detected in fecal
samples that are based on number of cycles or polymerases used.

When using the Bray-Curtis index, we observed that polymerases and
number of cycles had no affect on the bacterial community composition in
fecal samples (P-value = 1). There was also no interaction between
individuals, number of cycles, and polymerase used (P-value = 1). When
using PERMANOVA to test for community differences based on any of the
number of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03). Additionally, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 sequences
at 15 cycles. Within each respective 5-cycle increment comparison there
was no difference in Bray-Curtis index between the polymerases (P-value
\textgreater{} 0.05) {[}Figure 4B{]}. However, there was an overall
decrease in Bray-Curtis index by cycle comparison group (e.g.~the 15
cycle versus 20 cycle group compared to the 30 cycle versus 35 cycle
group) (P-value \textless{} 0.01) {[}Figure 4B{]}. Using a Dunn's
post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle
comparison groups had a higher Bray-Curtis index than the 30 cycle
versus 35 cycle comparison group (P-value \textless{} 0.05). In addition
to assessing community composition by Bray-Curtis index, we also
investigated differences in OTUs based on polymerase at each number of
cycle used and no significant differences could be found (P-value
\(\geqslant\) 0.01, corrected P-value = 1). Overall, these results
suggest that in fecal samples the number of cycles and polymerase used
minimally change the bacterial community composition.

@@

The small differences observed due to error rate and chimera prevalence
may be due to the actual DNA polymerase family being used within the PCR
mix. DNA polymerases from distinct families have different binding
affinities and error correction capacity (Ishino \& Ishino, 2014).
However, based on the observations within our study this is unlikely to
be the only contributor. Within our study the highest and lowest chimera
prevalence both belonged to a family A polymerase (Platinum and
Accuprime\textsuperscript{TM} respectively) (Ishino \& Ishino, 2014).
Additionally, based on the information supplied by the respective
manufacturers, the differences between the two PCR mixtures are not
immediately apparent. Both PCR mixtures contain a recombinant \emph{Taq}
DNA polymerase, a \emph{Pyrococcus} spp GB-D polymerase, and a platinum
\emph{Taq} antibody. Since it is not possible to know everything about
the mixture beyond what was willingly provided by the manufacturer, it
is possible that differences in how the recombinant \emph{Taq} was
generated or other compounds within the PCR mixture could be a
contributing factor for the differences in error rate and chimera
prevalence. Beyond the choice of the type of polymerase, there may be
other ways to reduce the affect of polymerase dependent error rates and
chimera prevalence on the downstream results.

Based on polymerase used, our observations generally found no difference
when using the Bray-Curtis index to measure the bacterial community.
Specifically, there was no difference in distance between successive
5-cycle increments within samples between the polymerases when using the
Bray-Curtis index {[}Figure 4{]}. One possible reason for this outcome
is that many of the OTUs generated by polymerase dependent error rates
and chimera prevalence are likely not highly abundant, allowing the
Bray-Curtis index to be able to successfully down-weight these OTUs
(Minchin, 1987). The choice of downstream diversity metric could be an
important consideration in helping to mitigate some of the observed
polymerase dependent differences in error rate and chimera prevalence.
Metrics that solely use presence/absence of OTUs (e.g.~Jaccard (Real \&
Vargas, 1996), richness) may be less robust to polymerase dependent
error rates and chimera prevalence. When choosing a distance metric,
careful consideration of the errors introduced from the PCR step of the
16S rRNA gene sequencing workflow need to be taken into account.

@@

\textbf{\emph{Mock community and human samples.}} The
ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
CA, USA) was used for mock communities and the bacterial component was
made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).
The theoretical abundance of 0.12 used as a comparison point for our
relative abundance was based on this equal genomic DNA composition of
the mock community. Fecal samples were obtained from 4 individuals who
were part of the Enterics Research Investigational Network (ERIN). The
processing and storage of these samples were previously published
(Seekatz et al., 2016). Other than confirmation that none of these
individuals had a \emph{Clostridium difficle} infection, clinical data
and other types of meta data were not utilized or accessed for this
study. All samples were extracted using the MOBIO\textsuperscript{TM}
PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA).

\textbf{\emph{PCR protocol.}} The five different high fidelity DNA
polymerases that were tested included AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The polymerases activation time was 2
minutes, unless a different activation was specified by the
manufacturer. The annealing and extension time for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time also were from a
previously published optimized protocol (Gohl et al., 2016). For
Phusion, the company defined activation and annealing times were used
while the extension time followed the Accuprime and Platinum settings.

The number of cycles in the PCR for mock and fecal samples started at 15
and increased by 5 up to 35 cycles, with amplicons used at each 5-step
increase for sequencing. The mock communities had 4 replicates at 15,
20, 25, and 35 cycles and 10 replicates for 30 cycles for all
polymerases (total samples n=130). No mock community sample had enough
PCR product at 15 cycles for adequate 16S rRNA gene sequencing. The PCR
of fecal DNA samples consisted of all 4 samples at 15, 20, 25, 30, and
35 cycles for each polymerase (total sample n=100).

\textbf{\emph{Sequence processing.}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis workflow.}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis. The
analysis of the mock community of each polymerase for GC-based
amplification differences, sequence error rate, number of sequences with
an error, base substitution, and numbers of chimeras before and after
chimera removal with VSEARCH was assessed. Additionally, the correlation
between the number of chimeras and the number of OTUs was also assessed.
The total number of OTUs, taxa differences, and Bray-Curtis indices were
analyzed for both the fecal and mock community samples. Finally, Random
Forest models were created to assess whether classification of
polymerase, number of cycles, or individual could be performed best
using 16S rRNA gene sequencing data. For the model used to classify
individuals, all samples were included from all number of cycles and
polymerases but only the individual label was used. Both number of
cycles and polymerase models included all samples but only the number of
cycles or polymerase label was used for each respective model.
Additionally, overlap between the most important OTUs to the three
models was assessed using mean decrease in accuracy (MDA).

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). High and low GC content was determined based on the median GC
percentage of either the V4 region or the average GC content of the
whole genome of the bacterial species used in the mock community.
Differences in the total number of OTUs were analyzed using an ANOVA
with a tukey post-hoc test. For the comparison of the number of OTUS in
fecal samples the data was normalized to each individual by cycle number
to account for the biological variation. Bray-Curtis distance matrices
were generated using mothur. The distance matrix data was analyzed using
PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
Kruskal-Wallis tests within R. The Random Forest models were run using
the caret package (v 6.0.78) (Jed Wing et al., 2017). A total of 100
10-fold CV runs on different 80/20 splits of the data were run to
generate a range of the Logloss value. The probability of a correct call
was obtained from this Logloss value by taking the negative natural
logarithm. For both error and chimera analysis, samples were tested
using Kruskal-Wallis with a Dunns post-hoc test. Where applicable,
correction for multiple comparison utilized the Benjamini-Hochberg
method (Benjamini \& Hochberg, 1995).

\textbf{\emph{Reproducible methods.}} The code and analysis for this
study can be found here
\url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}. The raw
sequences can be found on the SRA at the following accession number
SRP132931.
